---
title: "STA1040_EDA"
author: "Betelhem Getachew"
date: "2024-07-12"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Diabetes Dataset-Pima Indians

## About the dataset

This dataset is originally from the National Institute of Diabetes and
Digestive and Kidney Diseases. The objective of the dataset is to
diagnostically predict whether a patient has diabetes, based on certain
diagnostic measurements included in the dataset. Several constraints
were placed on the selection of these instances from a larger database.
In particular, all patients here are females at least 21 years old of
Pima Indian heritage.2 From the data set in the (.csv) File We can find
several variables, some of them are independent(several medical
predictor variables) and only one target dependent variable (Outcome).

## Information about dataset Attributes

-Pregnancies: To express the Number of pregnancies

-Glucose: To express the Glucose level in blood

-BloodPressure: To express the Blood pressure measurement

-SkinThickness: To express the thickness of the skin

-Insulin: To express the Insulin level in blood

-BMI: To express the Body mass index

-DiabetesPedigreeFunction: To express the Diabetes percentage -Age: To
express the age

-Outcome: To express the final result 1 is YES o is NO

## Installing and loading libraries

```{r warning=FALSE,message=FALSE}
#install.packages(c("tidyverse","VIM","visdat","DescTools"))
#install.packages("DescTools")
#install.packages("naniar")
#install.packages("fmsb")
#install.packages("moments")
#install.packages("caret")
#library(tidyverse)
#library(naniar)
#library(VIM)
#library(visdat)
#library(ggplot2)
#library(DescTools)
#install.packages("Metrics")
```

## Importing the dataset

```{r}
diabetes<-read.csv("C:\\Users\\Admin\\Desktop\\US2024\\STA\\diabetes.csv")
head(diabetes)
```

## Structure of the dataset

```{r}
str(diabetes)
```

## Data cleaning

### missing

```{r}
library(naniar)
any_na(diabetes)
```

#### Explanation

The function any_na() is part of the naniar package. If a missing value
exists in the dataset, it returns TRUE; otherwise, it returns FALSE.It
is evident from this example that certain values in our data set are
missing.

```{r}
diabetes <- diabetes %>% 
  mutate(across(c(Pregnancies, BloodPressure, Insulin, BMI, SkinThickness, DiabetesPedigreeFunction, Age),
                ~ replace(.x, .x == 0, NA)))

head(diabetes, 10)965699
```

### Handling missing values

```{r}
library(VIM)
diabetes<-kNN(diabetes,variable = colnames(diabetes),k=5,imp_var = FALSE)
head(diabetes)
```

#### Explanation

The kNN() function from the VIM package performs k-Nearest Neighbors
imputation on the sales_df dataframe, using 5 nearest neighbors to
estimate missing values. The kNN imputation method works as follows:
-For each missing value in the dataset, the algorithm identifies the k
nearest complete observations (based on the available variables). -The
missing value is then imputed using the mean or median of the
corresponding values from the k nearest neighbors.

### Data type Conversion

```{r}
diabetes$Outcome<-as.factor(diabetes$Outcome)
```

### Identifying outliers using the IQR

```{r}
identify_outliers <- function(diabetes) {
  # Initialize an empty list to store the results
  outlier_results <- list()
  
  # Loop through each numeric column
  for (col in names(diabetes)[sapply(diabetes, is.numeric)]) {
    # Calculate the quartiles and IQR
    q1 <- quantile(diabetes[[col]], 0.25)
    q3 <- quantile(diabetes[[col]], 0.75)
    iqr <- q3 - q1
    
    # Calculate the lower and upper bounds for potential outliers
    lower_bound <- q1 - 1.5 * iqr
    upper_bound <- q3 + 1.5 * iqr
    # Identify the outliers
    outliers <- diabetes[[col]] < lower_bound | diabetes[[col]] > upper_bound
    
    # Store the results
    outlier_results[[col]] <- list(
      outliers = which(outliers)
    )
  }
  
  return(outlier_results)
}
identify_outliers(diabetes)
```

## handling the outliers

```{r}
outlier_results <- identify_outliers(diabetes)

# Replace the outliers
for (col in names(outlier_results)) {
  # Get the outlier indices for the current column
  outlier_indices <- outlier_results[[col]]$outliers
  
  # Replace the outliers with the minimum and maximum values for that column
  min_value <- min(diabetes[[col]][-outlier_indices])
  max_value <- max(diabetes[[col]][-outlier_indices])
  diabetes[[col]][outlier_indices] <- pmin(pmax(diabetes[[col]][outlier_indices], min_value), max_value)
}
head(diabetes)
```

# Summary Statistics

## Measures of Central Tendency

## 1.Mean

### a.using formulas

Mean is a measure of central tendency which is the average of
observations. Manually it is calculated as follows.

```{r}
mean_age=sum(diabetes[["Age"]])/nrow(diabetes)
mean_age
```

### Explanation

Mean is calculated manually by summing up the observations and dividing
to the number of observations. So here we are adding up all ages and
dividing to the number of rows(observations).

### b.Using R's mean function

```{r}
mean(diabetes$Age)

```

### Explanation

It can be seen that we get the same result when we calculate the mean
manually and using R functions.

## 2. Median

The median is the value where 50% of the data is lower than it, and 50%
of the data is higher. We can calculate this by sorting all the data
points and taking the middle one.

### using formula

```{r}
sorted_insulin <- sort(diabetes$Insulin)
n <- length(sorted_insulin)

middle1 <- n / 2
middle2 <- middle1 + 1

median_insulin <- (sorted_insulin[middle1] + sorted_insulin[middle2]) / 2

median_insulin
```

### Explanation

since there are 786 observations the middle value is the average of the
two middle values.

### using R's median function

```{r}
median(diabetes$Insulin)
```

### Explanation

so the middle value of the insulin column is 126.5.

## 3. Mode

The mode is the most frequent value in the data

### using formula

```{r}
library(tidyverse)
mode_preg <- diabetes %>%
  count(Pregnancies) %>%
  arrange(desc(n)) %>%
  slice(1)
mode_preg
```

### using R's mode function

```{r}
library(DescTools)
Mode(diabetes $Pregnancies)
```

# Measures of Spread

## 1. Variance

measures the average distance from each data point to the data's mean.

### using formula

To calculate the variance, we start by calculating the distance between
each point and the mean, so we get one number for every data point.We
then square each distance and then add them all together.Finally, we
divide the sum of squared distances by the number of data points

```{r}
var_age<-sum((diabetes$Age-mean_age)^2)/nrow(diabetes)
var_age
```

### Using R's function

We can calculate the variance in one step using the var function.

```{r}
var(diabetes$Age)
```

## 2. Standard Deviation

The standard deviation is another measure of spread, calculated by
taking the square root of the variance. It can also be calculated using
the sd function.

### using formula

```{r}
std_age<-sqrt(var_age)
std_age
```

### using R's function

```{r}
sd(diabetes$Age)
```

so the deviation of the ages from the average age is almost 12 years.

## Range

Range is the difference between the maximum and minimum values of the
observation.

### using formula

```{r}
range_Glucose<-max(diabetes$Glucose)-min(diabetes$Glucose)
range_Glucose
```

### using R's function

```{r}
diff(range(diabetes$Glucose))
```

This values indicates that in the Diabetes dataset there is a difference
of 155 mg/dl of glucose level between the person who has the minimum
blood sugar and the person with maximum blood sugar.

## Quartiles

Quartiles split up the data into four equal parts. Here, we call the
quantile function to get the quartiles of the data.

### using formula

```{r}
sorted_Glucose <- sort(diabetes$Glucose)
# Number of observations
n <- length(sorted_Glucose)

# Calculate positions
pos_Q1_low <- n / 4
pos_Q1_high <- n / 4 + 1

pos_Q2_low <- n / 2
pos_Q2_high <- (n / 2) + 1

pos_Q3_low <- 3 * n / 4
pos_Q3_high <- (3 * n / 4 )+ 1

# Calculate quartiles
Q1 <- (sorted_Glucose[pos_Q1_low] + sorted_Glucose[pos_Q1_high]) / 2
Q2 <- (sorted_Glucose[pos_Q2_low] + sorted_Glucose[pos_Q2_high]) / 2
Q3 <- (sorted_Glucose[pos_Q3_low] + sorted_Glucose[pos_Q3_high]) / 2
Q1
Q2
Q3

```

### Using R's function

```{r}
quantile(diabetes$Glucose,c(0.25,0.5,0.75))
```

The above code indicates the followin -Q1:25% of the Glucose level
values in the dataset are below 99mg/dl -Q2:50% of the Glucose level
values in the dataset are below 117mg/dl which is also the median
-Q3:75% of the Glucose level values in the dataset are below 141mg/dl

## Deciles

Deciles are quantiles that divide a dataset into ten equal parts.

### using formula

Let us calculate from D1 up to D5

```{r}
sorted_Glucose <- sort(diabetes$Glucose)
# Number of observations
n <- length(sorted_Glucose)

# Calculate positions
pos_D1_low <- n / 10
pos_D1_high <- n / 10 + 1

pos_D2_low <- n / 5
pos_D2_high <- (n / 5) + 1

pos_D3_low <- 3 * n / 10
pos_D3_high <- (3 * n / 10 )+ 1

pos_D4_low <- 4 * n / 10
pos_D4_high <- (4 * n / 10 )+ 1

pos_D5_low <- 5 * n / 10
pos_D5_high <- (5 * n / 10 )+ 1


# Calculate quartiles
D1 <- (sorted_Glucose[pos_D1_low] + sorted_Glucose[pos_D1_high]) / 2
D2 <- (sorted_Glucose[pos_D2_low] + sorted_Glucose[pos_D2_high]) / 2
D3 <- (sorted_Glucose[pos_D3_low] + sorted_Glucose[pos_D3_high]) / 2
D4 <- (sorted_Glucose[pos_D4_low] + sorted_Glucose[pos_D4_high]) / 2
D5 <- (sorted_Glucose[pos_D5_low] + sorted_Glucose[pos_D5_high]) / 2
D1
D2
D3
D4
D5
```

### Using R's function

```{r}
quantile(diabetes$Glucose,probs=c(0.1,0.2,0.3,0.4,0.5))
```

Here it can be seen that 10% of the Glucose level values in the dataset
are below 86mg/dl.The other values are also interpreted the same way.

## Percentile

Percentile are quantiles that divide a dataset into ten equal parts.

### using formulas

Let us calculate from p1 up to p5

```{r}
sorted_Glucose <- sort(diabetes$Glucose)
# Number of observations
n <- length(sorted_Glucose)

# Calculate positions
pos_P1_low <- n / 100
pos_P1_high <- n / 100 + 1

pos_P2_low <- n / 50
pos_P2_high <- (n / 50) + 1

pos_P3_low <- 3 * n / 100
pos_P3_high <- (3 * n / 100 )+ 1

pos_P4_low <- 4 * n / 100
pos_P4_high <- (4 * n / 100 )+ 1

pos_P5_low <- 5 * n / 100
pos_P5_high <- (5 * n / 100 )+ 1


# Calculate quartiles
P1 <- (sorted_Glucose[pos_P1_low] + sorted_Glucose[pos_P1_high]) / 2
P2 <- (sorted_Glucose[pos_P2_low] + sorted_Glucose[pos_P2_high]) / 2
P3 <- (sorted_Glucose[pos_P3_low] + sorted_Glucose[pos_P3_high]) / 2
P4 <- (sorted_Glucose[pos_P4_low] + sorted_Glucose[pos_P4_high]) / 2
P5 <- (sorted_Glucose[pos_P5_low] + sorted_Glucose[pos_P5_high]) / 2
P1
P2
P3
P4
P5
```

### using R's function

```{r}
quantile(diabetes$Glucose,probs=c(0.01,0.02,0.03,0.04,0.05))
```

this means only 1% of the observations have a glucose level less than
67.67

## interquartile range

IQR is the difference between Q3 and Q1

### Using formula

```{r}
IQR<- Q3-Q1
IQR
```

### Using R's function

```{r}
IQR(diabetes$Glucose)
```

## Coefficient of variation(CV)

Coefficient of Variation (CV)is a measure of relative variability. CV is
a measure of variation that takes the scale of the dataset into account.
It is the ratio of standard deviation to mean. CV is more appropriate
when we want to compare the spread between two different sets of
observation that have different units of measurement.

**There is no specific function in R which is used to calculate CV.**

### Using formula

```{r}
cv_glucose<-sd(diabetes$Glucose)/mean(diabetes$Glucose)
cv_BP<-sd(diabetes$BloodPressure)/mean(diabetes$BloodPressure)
cv_glucose
cv_BP
```

#### Explanation

Glucose and blood pressure have different units of measurement, so their
variability(spread) can not compared using the standard deviation. CV is
a relative variability so now we can see that the value of "Glucose"
fluctuates more than "blood pressure" in this dataset.

## Mean Absolute Deviation(MAD)

The Mean Absolute Deviation (MAD) is a measure of the dispersion or
spread of a dataset. It represents the average of the absolute
differences between each data point and the mean of the dataset.

**There is no specific function for Mean Absolute Deviation in R.**

### using formula

```{r}
mad_skin<-sum(abs(diabetes$SkinThickness-mean(diabetes$SkinThickness)))/nrow(diabetes)
mad_skin
sd(diabetes$SkinThickness)
```

Unlike the standard deviation, which squares the deviations before
averaging, MAD uses the absolute value, making it less sensitive to
extreme values (outliers). As it can be seen sd is greater than MAD
because it is affected by extreme values.

## Semi Interquartile Range(SIQR)

The SIQR, being half of the IQR, represents the average distance from
the median to the quartiles (Q1 and Q3). Since the median is the
midpoint of the dataset, the SIQR gives a sense of how far, on average,
the data points within the central 50% of the distribution are from the
median.

### using formula

```{r}
siqr_glucose<-IQR/2
siqr_glucose
```

### using R's function

```{r}
library(fmsb)
SIQR(diabetes$Glucose)
```

This suggests that the data is spread out between the 25th percentile
(Q1) and the 75th percentile (Q3) by approximately 21 units. This means
that half of the data points fall within a range of 21 units from the
median.

# Measures of Shape of Distribution

## Skewness

Skewness is a measure of the extent to which a statistical distribution
is asymmetrical. It is a measure of the deviation of a distribution from
symmetry.

### calculating skewness using Quartiles

```{r}
skewness<-(Q3+Q1-2*Q2)/(Q3-Q1)
skewness
```

-Direction of skew: The positive value indicates that the distribution
is slightly right-skewed (or positively skewed). -Magnitude of skew: The
value is relatively close to 0, which suggests that the skewness is
mild.

### calculating skewness using Central Moments

```{r}
central_moment2<-sum((diabetes$Glucose-mean(diabetes$Glucose))^2)/nrow(diabetes)

central_moment3<-sum((diabetes$Glucose-mean(diabetes$Glucose))^3)/nrow(diabetes)

Skewness<-central_moment3/(central_moment2^(3/2))

Skewness
```

These two methods can often produce different results, sometimes
substantially so.this is because they have different sensitivities:

-The interquartile method is based on the positions of the quartiles and
is less sensitive to extreme values in the tails of the distribution.
-The central moments method takes into account all data points,
including extremes, making it more sensitive to outliers and the shape
of the distribution's tails.

### Using R's function

The widely used moments package in R uses central moments to calculate
skewness. The skewness() function in this package implements the
formula: skewness = m3 / (m2\^(3/2)) Where m3 is the third central
moment and m2 is the second central moment (variance).

```{r}
library(moments)
skewness(diabetes$Glucose)
```

Generally: -Values between -0.5 and 0.5 indicate approximately symmetric
data -Values between 0.5 and 1 (or -0.5 and -1) indicate moderate
skewness -Values greater than 1 (or less than -1) indicate high skewness
Based on this the distribution of the Glucose values is moderately
positively skewed.

This can be illustrated by using density plot.

```{r}
ggplot(diabetes,aes(x=Glucose))+
  geom_density(color="#4682B4")+
  labs(title="Density distribution of Glucose")
```

### Interpretation

-Shape: The distribution is right-skewed (positively skewed), with a
longer tail extending to the right. -Peak: The highest point of the
curve (mode) is around 90-100 mg/dL, indicating this is the most common
glucose level in the population. -Range: Glucose levels appear to range
from about 50 mg/dL to over 200 mg/dL. -Spread: There's considerable
spread in the data, with a significant portion of the population having
glucose levels between 75 and 150 mg/dL. -Tail: The long right tail
suggests some individuals have much higher glucose levels, possibly
indicating diabetic or pre-diabetic conditions. -Normal range: The bulk
of the distribution falls within what's typically considered the normal
fasting glucose range (70-100 mg/dL).

Here we can see that it is slightly skewed to the right.

## Kurtosis

peakedness or flatness of a distribution computed using 4th and 2nd
central moment. It measures the extent to which the data are distributed
in the tails versus the center of the distribution There are three types
of peakedness.

-   Leptokurtic: very peaked. The distribution has more outliers, with a
    higher peak and fatter tails. Kurtosis \> 3 , Excess kurtosis \>0

-   Platykurtic: relatively flat.The distribution has fewer outliers,
    with a lower peak and thinner tails. Kurtosis \< 3 , Excess kurtosis
    \< 0

-   Mesokurtic:in between.The distribution shape is similar to the
    normal distribution. Kurtosis = 3 , Excess kurtosis=0

### Using Formula

```{r}
central_moment2<-sum((diabetes$Glucose-mean(diabetes$Glucose))^2)/nrow(diabetes)

central_moment4<-sum((diabetes$Glucose-mean(diabetes$Glucose))^4)/nrow(diabetes)

kurtosis<- central_moment4/(central_moment2)^2
kurtosis


```

### using R's function

```{r}
kurtosis(diabetes$Glucose)
```

### Interpretation:

A kurtosis of 2.72 (or excess kurtosis of -0.28) indicates that the
distribution is slightly platykurtic. This means:

a)  The distribution has slightly lighter tails compared to a normal
    distribution.

b)  The peak is a bit flatter and broader than a normal distribution.

c)  There's a bit less probability in the tails and more probability in
    the shoulders of the distribution compared to a normal distribution.

d)  There are less outliers in this dataset

# Pearson coefficient of Correlation

Correlation is a statistical measure that describes the strength and
direction of the relationship between two variables.

Let us see the correlation between BMI and Glucose

### using Formula

```{r}
correlation<-sum((diabetes$BMI - mean(diabetes$BMI)) * (diabetes$Glucose - mean(diabetes$Glucose))) /  (sqrt(sum((diabetes$BMI- mean(diabetes$BMI))^2)) * sqrt(sum((diabetes$Glucose - mean(diabetes$Glucose))^2)))

correlation
```

### using R's function

```{r}
cor(diabetes$Glucose,diabetes$Insulin)
```

#### interpretation

Direction: The positive value (0.64) indicates a positive correlation.
This means that as Glucose tends to increase, insulin also tends to
increase.

Strength: 0.64 is considered a strong positive correlation. Generally,
correlations are interpreted as:

0.00 to 0.19: Very weak

0.20 to 0.39:Weak

0.40 to 0.59: Moderate

0.60 to 0.79: Strong

0.80 to 1.00: Very strong

### Visualizing correlation

```{r}
ggplot(diabetes,aes(Glucose,Insulin))+
  geom_point(color="#87CEEB")+
  geom_smooth(method="lm",se=FALSE)+
  labs(title="Insulin VS Glucose")
```

### Interpretation

With insulin on the y-axis, the upward trendline(best-fit line) suggests
a positive correlation between blood glucose and insulin. This means
that as blood sugar levels increase, the body appears to respond by
producing more insulin.

Physiological Correlation:

This positive correlation aligns with how the body regulates blood
sugar. Insulin is a hormone produced by the pancreas that helps cells
absorb glucose from the bloodstream. When blood sugar levels rise, the
pancreas releases more insulin to promote glucose uptake and lower blood
sugar levels.

# Simple Linear Regression

A simple linear model is a statistical method used to understand the
relationship between two continuous variables. It represents this
relationship using a straight line.

```{r}
library(caret)

set.seed(123)

# Creating a partition index
trainIndex <- createDataPartition(diabetes$Insulin, p = 0.7, list = FALSE)

# Split the data into training and testing sets
trainData <- diabetes[trainIndex, ]
testData <- diabetes[-trainIndex, ]

# Train the linear model
diabetes_model <- lm(Insulin ~ Glucose, data = trainData)
diabetes_model
```

### Description
This code uses the caret package to split the diabetes dataset into training (70%) and testing (30%) sets based on the Insulin variable, ensuring similar distributions in both sets. It then trains a linear regression model using the trainData set, with Insulin as the dependent variable and Glucose as the independent variable.

### Interpretation

-The linear model represents the equation of a best-fit line through the
data points of glucose vs insulin levels in the "diabetes" dataset.

-The slope (1.64) in this model signifies the change in insulin levels
for every one unit increase in glucose. In this case, for every 1 mg/dL
increase in glucose, the model predicts an average increase of 1.64
units in insulin.

Reconciling the Findings:

The positive correlation (0.64) and the positive slope (1.64) are
actually consistent with each other. They both indicate that when
glucose levels go up, insulin levels tend to go up as well. This aligns
with the body's natural response to regulate blood sugar.
### Making predictions

```{r}
# Make predictions on the test data
predictions <- predict(diabetes_model, newdata = testData)

head(predictions)
```
The values are the predicted insulin levels using the diabetes model.

### summary of the model

```{r}
summary(diabetes_model)
```

### Interpretation
The linear regression model predicting Insulin levels based on Glucose levels shows that the intercept is -37.90585, indicating the baseline insulin level when glucose is zero. The coefficient for glucose is 1.43357, which is highly significant (p-value < 2e-16), suggesting that for every unit increase in glucose, insulin increases by 1.43357 units. The residuals range from -206.138 to 191.380, with a median of -8.998 and a residual standard error of 54.48. The model explains approximately 40.15% of the variance in insulin levels (Multiple R-squared: 0.4015, Adjusted R-squared: 0.4004). The F-statistic is 360.2 with a highly significant p-value (< 2.2e-16), indicating that the model is a good fit for the data.

### Metrics of Linear Regression

```{r}
library(Metrics)
mse<-mse(testData$Insulin,predictions)
mse
rmse<-sqrt(mse)
rmse
```
#### Interpretation

-Mean Squared Error (MSE)
Interpretation: The MSE represents the average of the squared differences between the predicted values and the actual values. A lower MSE indicates that the model's predictions are closer to the actual values. In this case, the MSE of 2981.098 suggests that, on average, the squared difference between the predicted and actual values is 2981.098. However, since MSE is in the squared unit of the dependent variable, it can be hard to interpret directly.

-Root Mean Squared Error (RMSE)
Interpretation: The RMSE is the square root of the MSE and provides a measure of the average magnitude of the prediction errors. It is in the same unit as the dependent variable, making it more interpretable. An RMSE of 54.59943 means that, on average, the model's predictions are off by about 54.6 units from the actual values. The lower the RMSE, the better the model's performance.
# Visualization

Data visualization is an essential skill for data scientists. It
combines statistics and design in meaningful and appropriate ways. On
the one hand, data vis is a form of graphical data analysis, emphasizing
accurate representation and interpretation of data. On the other hand,
data vis relies on good design choices, not only to make our plots
attractive, but to also aid both the understanding and communication of
results. On top of that, there is an element of creativity, since at
it's heart, data vis is a form of visual communication.

## bar Chart

```{r}
ggplot(diabetes,aes(Outcome))+
  geom_bar(position="dodge",fill="#87CEEB")+
   labs(title="Count of Diabetic and Non-Diabetic Individuals", 
       x="out come of diagnosis", 
       y="Count") +
  scale_x_discrete(labels = c("0" = "Non-diabetic", "1" = "Diabetic")) +
  theme(axis.title=element_text(size=12,color="blue"),plot.title=element_text(size=12,color="blue"),line=element_blank(),rect=element_blank())
```

### Interpretation

From this bar graph it can be seen that 500 individuals are not diabetic
and around 265 individuals are diabetic.

## Histogram

A histogram is a special type of bar plot that shows the binned
distribution of a continuous variable.

```{r}
ggplot(diabetes,aes(x=BMI))+
  geom_histogram(binwidth=1,fill="#4682B4")+
  facet_wrap(~ Outcome)+
  labs(title="Distribution of BMI")+
  facet_wrap(~ Outcome, labeller = as_labeller(c(`0` = "Non-diabetic", `1` = "Diabetic"))) +
  theme(axis.title=element_text(size=12,color="#4682B4"),plot.title=element_text(size=12,color="#4682B4"))
```

### Interpretation

Here are some observations about the distribution of BMI in the two
groups:

The non-diabetic group appears to have a wider distribution of BMI than
the diabetic group. This means that the non-diabetic group has a larger
range of BMIs, with some people having very low BMIs and others having
very high BMIs. The diabetic group, on the other hand, seems to be more
concentrated in the overweight and obese BMI ranges (30-40 and 40-50).
The non-diabetic group appears to have a peak density around a BMI of
25, which is considered to be the normal weight range. The diabetic
group, on the other hand, appears to have a peak density around a BMI of
30, which is considered to be the overweight range.

## density plot

Density plots are similar to histograms, but instead of showing the
count of data points in each bin, they show the probability
density.Density plot represents the shape of the histogram using smooth
line. It gives a bigger picture of the shape of the distribution

```{r}
ggplot(diabetes,aes(BMI))+
  geom_density(binwidth=1,fill="#4169E1")+
  facet_wrap(~ Outcome, labeller = as_labeller(c(`0` = "Non-diabetic", `1` = "Diabetic"))) +
  labs(title="Density plot of BMI in diabetic and nondiabetic")+
  theme(axis.title=element_text(color="#4169E1"),plot.title=element_text(color="#4169E1"))
  
```

### Interpretation

-People without diabetes show a wider spread of body mass index (BMI)
compared to those with diabetes. This means there's a larger variety of
BMIs in the non-diabetic group, with some individuals being very
underweight and others very overweight. In contrast, the diabetic group
seems to be clustered around higher BMIs, particularly in the overweight
and obese ranges (30-40 and 40-50).

-Looking at the peak density, the non-diabetic group tends to have a
healthy BMI around 25, which is considered the normal weight range. On
the other hand, the peak density for the diabetic group falls around 30,
which is already in the overweight category.

## Cummulative frequency plot(Ogive)

```{r}
#  ogive plot for the age column with a bin width of 10
ggplot(diabetes, aes(x = Age,color=Outcome)) +
  stat_bin(aes(y = cumsum(..count..)), binwidth = 10, geom = "line", size = 1) +
  labs(
    title = "Ogive (Cumulative Frequency Polygon) of Age",
    x = "Age",
    y = "Cumulative Frequency"
  ) +
  scale_color_manual(values = c("blue", "#4682B4"), 
                     labels = c("Not-Diabetic", "Diabetic"),
                     name = "Diabetes Status")+
  theme(axis.title=element_text(color="#4169E1"),plot.title =element_text(color="#4169E1"))
  
```

### Interpretation

-Steep Slope: A steeper initial slope on the ogive indicates a rapid
rise in the number of people with diabetes within a specific age range.
This could suggest factors like hormonal changes, increased risk of
pre-diabetes progressing to diabetes, or environmental triggers
impacting this age group more significantly.

-Plateau or Gradual Increase: A plateau or gradual increase in the curve
later in life might indicate a saturation of at-risk individuals
developing diabetes, or potentially a shift in risk factors impacting
older age groups differently.

## Frequency Polygon

A frequency polygon is a graphical representation of the distribution of
a dataset. It is similar to a histogram but uses a line graph instead of
bars to represent the frequency of data points within specified
intervals (bins). The frequency polygon provides a clearer picture of
the shape of the data distribution and is particularly useful for
comparing multiple distributions on the same graph.

```{r}
ggplot(diabetes,aes(x=Glucose,color=Outcome))+
  geom_freqpoly()+
  labs(title="Frequency polygon of Glucose level for diabetic and non diabetic")+
  scale_color_manual(values = c("#00008B", "#1E90FF"), 
                     labels = c("Not-Diabetic", "Diabetic"),
                     name = "Diabetes Status")+
theme(axis.title=element_text(color="#1E90FF"),plot.title =element_text(color="#1E90FF"))
```

### Interpretation

-Diabetic (Light Blue Line): The distribution for diabetic individuals
peaks around 150-180 glucose levels. This distribution is wider and has
multiple peaks, indicating more variability in glucose levels among
diabetic individuals.

-Non-Diabetic (Dark Blue Line): The distribution for non-diabetic
individuals peaks around 100-120 glucose levels. This distribution is
narrower with a single, prominent peak, suggesting less variability in
glucose levels among non-diabetic individuals.

-Comparison: Diabetic individuals generally have higher glucose levels
compared to non-diabetic individuals. The distribution for diabetic
individuals is more spread out with higher glucose levels, while
non-diabetic individuals have a more concentrated distribution around
lower glucose levels.

-Skewness: The non-diabetic distribution is slightly right-skewed. The
diabetic distribution appears more irregular with peaks at various
glucose levels, indicating a mix of moderate to high glucose levels.

## Scatter plot

```{r}
ggplot(diabetes,aes(x=Insulin,y=Glucose))+
  geom_point(position="jitter",color="#87CEEB")+
  labs(title="Glucose level VS Insulin level")+
  theme(axis.title=element_text(color="#1E90FF"),plot.title =element_text(color="#1E90FF"))
```

### Interpretation

-Correlation: There appears to be a positive correlation between insulin
and glucose levels. As insulin levels increase, there's a general trend
of glucose levels also increasing.

-Spread: The data points are widely spread, indicating significant
variability in the relationship between insulin and glucose.

-Concentration: The majority of data points are clustered in the lower
left to middle part of the graph, suggesting that most individuals have
moderate levels of both insulin and glucose.

-Outliers: There are several outliers, particularly on the right side of
the graph, showing some individuals with very high insulin levels.

-Non-linear relationship: The relationship doesn't appear to be strictly
linear. There seems to be more variation in glucose levels at higher
insulin levels.

-Range: Insulin levels range from near 0 to about 300, while glucose
levels range from about 50 to 200.

## Stem and Leaf Plot

A stem-and-leaf plot is a graphical method used to display quantitative
data in a way that retains the original data values. It is particularly
useful for small to moderate-sized data sets and provides a quick way to
assess the shape of the distribution, identify the median, and detect
any outliers.

-   Stem: Represents the leading digits of each data value.

-   Leaf: Represents the trailing digit(s) of each data value.

```{r}
filtered_diabetes <- diabetes %>% filter(Outcome=="1")

stem(filtered_diabetes$Glucose,scale=1)

```

### Interpretation 

-General Shape: The plot shows
how the data is distributed across different ranges. Most of the data
points are concentrated in the stems 10 through 14.

Spread: The glucose values range from the 70s to the 190s. There are no
values below 70 or above 199 in this dataset.

-Symmetry: The distribution is somewhat symmetrical, but there is a
slight right skew (more values towards the higher end). There are more
values concentrated in the higher stems (10 to 14) compared to the lower
stems (7 to 9).

## Box plot

```{r}
par(mfrow = c(4, 2), mar = c(3, 3, 2, 1)) 
for (col in names(diabetes)[sapply(diabetes, is.numeric)]) {
  boxplot(diabetes[[col]], 
          main = paste("Boxplot of", col), 
          ylab = "Values",cex.main = 1.2,cex.lab = 1.1)
}
```

### Interpretation

-Pregnancies: Shows a right-skewed distribution with most values
clustered at the lower end. -Glucose: Appears relatively symmetrical
with a few outliers on the lower end.

-Blood Pressure: Fairly symmetrical distribution with some outliers on
both ends. -Skin Thickness: Somewhat symmetrical with a few outliers on
the lower end.

-Insulin: Highly right-skewed distribution with many outliers on the
upper end.

-BMI (Body Mass Index): Slightly right-skewed with a few outliers on
both ends.

-DiabetesPedigreeFunction: Highly right-skewed with many outliers on the
upper end.

-Age: Right-skewed distribution with several outliers on the upper end.

we can also plot the distribution of Glucose level among diabetics and
non diabetics.

## Boxplot-2

```{r}
ggplot(diabetes,aes(x=Outcome,y=Glucose))+
  geom_boxplot(color="#1E90FF")+
  labs(title="Boxplot of Glucose for Diabetics and Non-diabetics",x="Diagnosis Outcome")+
  scale_x_discrete(labels = c("0" = "Non-diabetic", "1" = "Diabetic"))+
  theme(axis.title=element_text(color="#1E90FF"),plot.title =element_text(color="#1E90FF"))
```

### Interpretation

**Why are there outliers even though outliers where handled at the
begining?** When you plot each variable (including glucose) for the
entire dataset, the calculation of outliers is based on the overall
distribution of that variable. However, when you subset the data into
diabetic and non-diabetic groups, the distribution of glucose levels
within each group might differ significantly from the overall
distribution, leading to different sets of outliers.

**Distribution**

-Median: The median glucose level for diabetics is higher than for
non-diabetics, as shown by the horizontal line inside each box.

-Spread: Diabetics have a wider range of glucose values, indicated by
the taller box and longer whiskers.

-Outliers: Both groups have outliers, represented by individual points
beyond the whiskers. Non-diabetics have several high outliers, while
diabetics have a few low outliers.

-Quartiles: The interquartile range (box size) is larger for diabetics,
suggesting more variability in glucose levels.

-Overall range: Diabetics show a higher maximum glucose level and a
slightly higher minimum level compared to non-diabetics. -Distribution:
The non-diabetic box appears more symmetrical, while the diabetic box is
slightly skewed upwards, indicating a tendency towards higher glucose
levels.

## Pie chart

```{r}
outcome_counts <- table(diabetes$Outcome)
percentages <- round(100 * outcome_counts / sum(outcome_counts), 1)

# Create labels with percentages
labels <- paste(c("Not Diabetics", "Diabetics"), "\n", percentages, "%", sep = "")

# Plot the pie chart with percentages
pie(outcome_counts, 
    labels = labels,
    main = "Distribution of Diabetes Outcomes",
    col = c("lightblue", "darkblue"))
```

From the above plot it can be seen that 65.1% of the individuals in this
data set are not diabetics and 34.9% are diabetics.

## Line Plot Using Amazon's Stock Market Data

```{r}
amazon<-read.csv("C:\\Users\\Admin\\Desktop\\US2024\\STA\\amazon_stock_data.csv")
head(amazon)
```
## Cleaning the data

```{r}
any_na(amazon)
```
## Time Series Plot of the closing price of Amazon
```{r}
amazon$Date <- as.Date(amazon$Date) 
ggplot(amazon,aes(x=Date,y=Close))+
  geom_line(color="darkblue")+
  labs(x="Year",y="Closing Price of Amazon",title="Amazon Stock Price Over Time")+
  theme(axis.title=element_text(color="darkblue"),plot.title=element_text(color="darkblue"))
```

### Interpretations

-There is a generally upward trend in the stock price over time, indicating that Amazon's stock price has been increasing over the past several years.

-There are also some fluctuations in the stock price over time, with periods of both increase and decrease.
